\vspace{-15mm}\fontsize{24pt}{10pt}\selectfont\textbf{Assessing economic inequality with tax data - Switzerland from 1945 to 2010}

### 2014-07-02 paper/subdocuments/methods.Rnw
fixed formulae

The RD-framework is based on the concept of a ``relative distribution'', a transformation of the data from two distributions into a single distribution that contains all of the information necessary for scale-invariant comparison. This allows the make distribution differences ``visual'' in an elegant way and it is also a base for summary statistics, which are more sensitive to detailed theoretical hypotheses in contrast to other measures like the gini-coefficient or top income-shares, which inform either about the whole population or only about the top part of the distribution.\\

% Probability Density function as a base for the RD
The goal of RD is to study the differences between two distributions. A common example could be the income distribution for man and women. Subject of the comparison can also be two distributions describing the same population, but stemming from two different data sources like survey or tax data or even comparison of the same source/population but for different time points, like the income distribution out of tax data for Switzerland today compared to an earlier time point. To describe what we gone compare, we start with defining the two distributions. On represents the reference population (\Y_{0}) and the other the comparison population \Y. Y represents our measure of interest (income). A first visual approach is to compare the two probability density functions (PDF). The PDF is a function (fy) which describes the distribution of probability over the outcome set and is defined for all possible values of y. Also this function integrates to 1, which means that the sum of all probabilities over all possible values is 100\%. Out of the comparison of the PDF, it is possible to see, which values of x are more and less probable. This allows allready to spot distributional differences over the whole scale of Y visualy. \\

The PDF can be characterized by its cumulative distribution function (CDF). The CDF can be formulated as F(y), which represents the cumulative probability at point x, which is the probability that a randomly chosen value is less than or equal to y. The relative distribution of \Y to \Y_{0} is then defined as

@@

R is obtained from \Y by transforming it by the CDF for \Y_{0}, \F_{0}. R measures therefore the relative rank of Y compared to Yo.

@@

We can calculate the Probability Density Function (g(r)) of RD, where \r represents the proportion of values and F_{0}^{-1}(r) is the inverse cumulative distribution function, also called the quantile function. g(r) can be interpreted as a density Ratio, which is defined as the ratio of these two quantities of every percentile of the reference distribution [0,1]. With a complete overlap of both distributions the Probability Density Function of the RD is 1 at every point of the PDF. On the other hand, values higher than 1 represent higher probabilities in the comparison distribution than in the references distribution at this specific point and values lower than 1 respectively represent lower probabilities. It is a proper PDF in the sense that it integrates to 1 over the unit interval.\\

% Median and shape differences
What we get through the above transformation of two distributions is the overall relative probability density. But Differences between distributions can be divided into two basic components: changes in location and changes in shape. If the comparative distribution is a simple location-shifted version of the reference distribution, then the difference between the two distributions can be parsimoniously summarized by this shift. Differences that remain after a location adjustment are differences in ``shape'' (scale, skew and other distributional characteristics). When both types of shifts are operating, or when factors other than scale are changing in the shape component, we need a way to separate out the various effects. If we want to identify the effect of a location shift and separate it from other changes in the distribution, it is necessary to specify what scale this shift operates on. It is possible to adjust distribution by any measure of central tendency. We choose here a median location adjustment because the median is a natural, robust and scale invariant unit of measurement. Because our interest lies in analyzing distributional differences concerning the degree of inequality, we will focus in the results section on shape differences and look therefore at the relative distribution after the distributions are adjusted for location differences.

% RD based Summary Meausres
Distributional polarization is of particular interest in the study of inequality. However, common inequality indicator (for example Gini or Theil’s index) are not designed to distinguish between growth in the upper and lower tails. Even if the measures register increasing inequality over time, one cannot distinguish a polarization of the distribution (increases in both tails) from upgrading (increases in the upper tail), downgrading (increases in lower tail). The polarization index plays the same role as the difference in Gini-indices, but it can be decomposed to compare the growth in the upper and lower tails. Because it is based on the relative distribution it provides a simple link between what is observed in the graphical display and what is measured by the numerical summary. \\

The median relative polarization index (MRP) is then the mean absolute deviation around the median of the location-adjusted relative distribution, scaled to produce an index that varies between -1 and 1. Given the scaling, a value of zero represents no differences in distributional shape; positive values represent more polarization (increases in the tails of the distribution); and negative values represent less polarization (convergence towards the center of the distribution). The measure catches only differences in distributional shape (not location). And it has several interesting features. MRP can be interpreted in terms of a proportional shift of mass in the distribution from more central to less central values. A value of 0.1, for example, is equivalent to a 10\% population shift from the center of the distribution to the upper and lower quartiles and the MRP is decomposable along the scale of y. This makes it possible to compare the contribution of each section of the distribution to the overall polarization. A natural decomposition is the contributions made by components above and below the median (of g(r)).

@@

<<zero_descriptives>>=
library(ggplot2)
ggplot(df[df$steuerperiode>=1995&df$kanton=="CH",], aes(x=steuerperiode,y=zeros))+geom_line()+facet_wrap(~kanton)+theme_bw()
#ggplot(df[df$steuerperiode>=1995,], aes(x=steuerperiode,y=zeros))+geom_line()+facet_wrap(~kanton)+geom_line(aes(y=G_steink),color="blue")+theme_bw()
@

The number of zeros increased slightly. Different interpretations seems possible. The increased share of zeros could be driven by the FTA's inflation adjustments or an increase in social deductions. Another plausible explanation is the increasing number of unmarried couples filling in two separate tax forms instead of one so we observe two ``small'' incomes instead of one ``large'' income, thus inflating the number of people falling below the taxation threshold.

@@

By adding the zeros back as a separate group we need to make an assumption about the income distribution within that group. We know these cases had incomes between zero and the threshold (usually around 15.000 CHF for the period 1995-2010). In figure WITHWITHOUTZEROSTIMESERIES we show Gini coefficients calculated under the assumption of the zeros being zero. This is problematic for two reasons: first we know these units might earn more (up to the threshold) so Ginis calculated on this assumptions will be too high. Second we assume the zero group to be perfectly homogenous which on the contrary has a lowering effect on the Gini coefficient. Therefore we must at least assume a distribution for the zeros that has some variance and ranges from zero to the threshold. We decided to use the most simple approach fulfilling both criteria which is to interpolate a uniform distribution between zero and each tax periods threshold.

@@

<<corrected_gini, dev='bmp'>>=
df$J<-factor(df$steuerperiode)
fit <- lm(G_steink~kanton+kanton:zeros+J+zeros:J,data=df)
#summary(fit)
SmoothCoefficientPlot <- function(models, modelnames = "", removeintercept = FALSE){
  # models must be a list()

@@

Figure COEFPLOT can be read as follows: A positive coefficient (e.g. Schwyz) states that we measure higher Gini coefficients in periods with many zeros. We can derive, that the distribution of incomes is more skewed for high incomes than for low incomes. Simply speaking, the contrast between low and middle class is less pronounced than the contrast between middle and top class. One possible explanation would be that incomes stem from two different distributions (populations): 1) local people of Schwyz who follow a less skewed distribution and 2) particularly rich people who moved to Schwyz (to avoid taxes).

>>>>

The RD-framework is based on the concept of a ``relative distribution'', a transformation of the data from two distributions into a single distribution that contains all of the information necessary for scale-invariant comparison. This allows to make distribution differences ``visual'' in an elegant way and it is also a base for summary statistics, which are more sensitive to detailed theoretical hypotheses in contrast to other measures like the gini-coefficient or top income-shares, which inform either about the whole population or only about the top part of the distribution.\\

% Probability Density function as a base for the RD
The goal of RD is to study the differences between two distributions. A common example could be the income distribution for man and women. Subject of the comparison can also be two distributions describing the same population, but stemming from two different data sources like survey or tax data or even comparison of the same source/population but for different time points, like the income distribution out of tax data for Switzerland today compared to an earlier time point. To describe what we are going to compare, we start with defining the two distributions. One represents the reference population $Y_{0}$ and the other the comparison population $Y$. $Y$ represents our measure of interest (income). A first visual approach is to compare the two probability density functions (PDF). The PDF is a function $f(y)$ which describes the distribution of probability over the outcome set and is defined for all possible values of $y$. Also this function integrates to 1, which means that the sum of all probabilities over all possible values is 100\%. Out of the comparison of the PDF, it is possible to see, which values of $x$ are more and less probable. This allows already to spot distributional differences over the whole scale of $Y$ visually. \\

The PDF can be characterized by its cumulative distribution function (CDF). The CDF can be formulated as $F(y)$, which represents the cumulative probability at point $x$, which is the probability that a randomly chosen value is less than or equal to $y$. The relative distribution of $Y$ to $Y_{0}$ is then defined as

@@

R is obtained from $Y$ by transforming it by the CDF for $Y_{0}$, $\F_{0}$. $R$ therefore measures the relative rank of $Y$ compared to $Y_{0}$.

@@

We can calculate the Probability Density Function $g(r)$ of RD, where $r$ represents the proportion of values and $F_{0}^{-1}(r)$ is the inverse cumulative distribution function, also called the quantile function. $g(r)$ can be interpreted as a density ratio, which is defined as the ratio of these two quantities evaluated at every percentile of the reference distribution [0,1]. With a complete overlap of both distributions the probability density function of the RD is 1 at every point of the PDF. On the other hand, values higher than 1 represent higher probabilities in the comparison distribution than in the references distribution at this specific point and values lower than 1 respectively represent lower probabilities. It is a proper PDF in the sense that it integrates to 1 over the unit interval.\\

% Median and shape differences
What we get through the above transformation of two distributions is the overall relative probability density. But differences between distributions can be divided into two basic components: changes in location and changes in shape. If the comparative distribution is a simple location-shifted version of the reference distribution, then the difference between the two distributions can be parsimoniously summarized by this shift. Differences that remain after a location adjustment are differences in ``shape'' (scale, skewness and other distributional characteristics). When both types of shifts are operating, or when factors other than scale are changing in the shape component, we need a way to separate out the various effects. If we want to identify the effect of a location shift and separate it from other changes in the distribution, it is necessary to specify what scale this shift operates on. It is possible to adjust distributions by any measure of central tendency. Here we choose a median location adjustment because the median is a natural, robust and scale invariant unit of measurement. Because our interest lies in analyzing distributional differences concerning the degree of inequality, we will focus in the results section on shape differences and look therefore at the relative distribution after the distributions are adjusted for location differences.

% RD based Summary Meausres
Distributional polarization is of particular interest in the study of inequality. However, common inequality indicators (for example Gini or Theil’s index) are not designed to distinguish between growth in the upper and lower tails. Even if the measures register increasing inequality over time, one cannot distinguish a polarization of the distribution (increases in both tails) from upgrading (increases in the upper tail) or downgrading (increases in lower tail). The polarization index plays the same role as the difference in Gini-indices, but it can be decomposed to compare the growth in the upper and lower tails. Because it is based on the relative distribution it provides a simple link between what is observed in the graphical display and what is measured by the numerical summary. \\

The median relative polarization index (MRP) is then the mean absolute deviation around the median of the location-adjusted relative distribution, scaled to produce an index that varies between -1 and 1. Given the scaling, a value of zero represents no differences in distributional shape; positive values represent more polarization (increases in the tails of the distribution); and negative values represent less polarization (convergence towards the center of the distribution). The measure catches only differences in distributional shape (not location). And it has several interesting features. MRP can be interpreted in terms of a proportional shift of mass in the distribution from more central to less central values. A value of 0.1, for example, is equivalent to a 10\% population shift from the center of the distribution to the upper and lower quartiles and the MRP is decomposable along the scale of $y$. This makes it possible to compare the contribution of each section of the distribution to the overall polarization. A natural decomposition is the contributions made by components above and below the median (of $g(r)$).

@@

<<zero_descriptives, fig.cap='',fig.pos='H'>>=
library(ggplot2)
ggplot(df[df$steuerperiode>=1995&df$kanton=="CH",], aes(x=steuerperiode,y=zeros))+geom_line()+theme_bw()+xlab("tax period")+ylab("share of non-taxed")
#ggplot(df[df$steuerperiode>=1995,], aes(x=steuerperiode,y=zeros))+geom_line()+facet_wrap(~kanton)+geom_line(aes(y=G_steink),color="blue")+theme_bw()
@

The number of zeros increased slightly according to figure \ref{fig:zero_descriptives}. Different interpretations seems possible. The increased share of zeros could be driven by the FTA's inflation adjustments or an increase in social deductions. Another plausible explanation is the increasing number of unmarried couples filling in two separate tax forms instead of one so we observe two ``small'' incomes instead of one ``large'' income, thus inflating the number of people falling below the taxation threshold.

@@

By adding the zeros back as a separate group we need to make an assumption about the income distribution within that group. We know these cases had incomes between zero and the threshold (usually around 15.000 CHF for the period 1995-2010). In figure \ref{fig:with_without_zeros_Gini} we show Gini coefficients calculated under the assumption of the zeros being zero. This is problematic for two reasons: first we know these units might earn more (up to the threshold) so Ginis calculated on this assumptions will be too high. Second we assume the zero group to be perfectly homogenous which on the contrary has a lowering effect on the Gini coefficient. Therefore we must at least assume a distribution for the zeros that has some variance and ranges from zero to the threshold. We decided to use the most simple approach fulfilling both criteria which is to interpolate a uniform distribution between zero and each tax periods threshold.

@@

<<corrected_gini, dev='bmp', fig.cap='',fig.pos='H'>>=
df$J<-factor(df$steuerperiode)
fit <- lm(G_steink~kanton+kanton:zeros+J+zeros:J,data=df)
#summary(fit)
SmoothCoefficientPlot <- function(models, modelnames = "", removeintercept = FALSE){
  # models must be a list()

@@

Figure \ref{fig:corrected_gini} can be read as follows: A positive coefficient (e.g. Schwyz) states that we measure higher Gini coefficients in periods with many zeros. We can derive, that the distribution of incomes is more skewed for high incomes than for low incomes. Simply speaking, the contrast between low and middle class is less pronounced than the contrast between middle and top class. One possible explanation would be that incomes stem from two different distributions (populations): 1) local people of Schwyz who follow a less skewed distribution and 2) particularly rich people who moved to Schwyz (to avoid taxes).

### 2014-07-02 paper/subdocuments/results.Rnw
fixed formulae

gA0 <- reldist(y=bd.2010.ind, yo=bd.0.2010.ind,
               smooth=0.4, ci=FALSE,
               show="residual",
               entropy=NA,
               bar=TRUE, quiet=FALSE,
               ylim=c(0.5,2), ylab="",
               yolabs=seq(-1,3,by=0.5),
               discrete=FALSE,
               xlab="Proportion of 2010 Distribution (including non-taxed)")
title(main=paste("Effect of different shape"))
abline(h=1,lty=2)
@

>>>>

gA0 <- reldist(y=bd.2010.ind, yo=bd.0.2010.ind,
               smooth=0.4, ci=FALSE,
               show="residual",
               bar=TRUE, quiet=FALSE,
               ylim=c(0.5,2), ylab="",
               discrete=FALSE,
               xlab="Proportion of 2010 Distribution (including non-taxed)")
title(main=paste("Effect of different shape"))
abline(h=1,lty=2)
@
